Energy Clustering

### 2017-10-03 energy_clustering/draft/energy_clustering.tex
Finished 1st revision

\newcommand\Energy{\mathcal{E}}
\newcommand\EnergyH{\mathcal{E}^{H}}
\newcommand\EnergyL{\mathcal{E}^{L}}
\newcommand\EnergyOne{\mathcal{E}^{1D}}
\newcommand\E{\mathbb{E}}
\newcommand\kk{K}
\newcommand\kkk{h}
\newcommand\Hk{{\mathcal{H}}_{\kk}}
\newcommand\HH{\mathcal{H}}
\newcommand\C{{\mathcal{C}}}
\newcommand\tC{{\widetilde{\C}}}
\newcommand\OO{{\mathcal{O}}}
\newcommand\Zt{Y}
\newcommand{\Ind}[1]{\mathbbm{1}_{#1}}
\newcommand\e{e}
\newcommand\om{\omega}

@@

\begin{abstract}
Energy statistics was proposed by Sz\' ekely in the 80's inspired by the 
Newtonian gravitational potential from classical mechanics, and it provides 
a hypothesis test for equality of distributions. Energy statistics was 
further generalized to data living metric spaces of strong negative type, 
and more recently a connection with reproducing kernel Hilbert spaces (RKHS) 
was proposed. Here we consider the problem of clustering data from an 
energy statistics 
theory perspective, called energy clustering for short, by providing a 
precise mathematical formulation yielding a quadratically constrained 
quadratic program (QCQP) in the associated RKHS, 
which we show to be equivalent to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of kernel $k$-means 
from energy statistics.
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, making connection to graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative algorithm based 
on Hartigan's method, which in this case has the same computational cost 
as kernel $k$-means algorithm, based on Lloyd's method, but usually 
with better clustering quality. 
We provide carefully designed numerical experiments showing the superiority 
of the proposed method compared to kernel $k$-means, standard $k$-means, 
and gaussian mixture models in a variety of settings.
\end{abstract}

@@

@@

% @gui: algorithms don't make assumptions, theorems do.  also, algorithms cannot be parametric or non-parametric, that is a property only of the statistical model, or an estimator.

% @gui: k-means is NOT a special case of GMM, GMM is a soft clustering (not fuzzy, in stats and machine learning we don't say "fuzzy"), where as k-means is a hard clustering.  k-means is certainly closely related to GMM though.

% @gui:  don't know what "clear resolution of cluster centers" adds to this?

% @gui: algorithms cluster, doesn't matter how complicated the data are.  so, what do you actually mean here?  perform well empirically? what does "complicated" actually mean?

% @gui: the words "this" and "it" are often vague and ambiguous, be specific/precise

% @gui: not sure what you mean here, in kernel k-means, we don't "optimize" over any parameters typically.  sometimes we select hyperparameters of kernels via cross-validation, is that what you mean?

% @gui: not sure what you mean "only one possible optimization problem".  we can reformulate it many different ways? also, below, "this" is ambiguous.

% @gui: to me, we did not make explicit one of the most important points.   in this paper, we make clear exactly when we expect energy clustering to perform better than k-means/GMM, and when we expect it to perform worse (empirically).

% @gui: i find ToC paragraphs useless.

% @jovo: corrected all of these points, and also re-wrote some other stuff. Please check. About ToC, I agree and wouldn't mind remove it, however it is a tradition that most referees like to see. In this case I included some important information about our work too, and it makes a transition between introduction and the remaining of the paper. If you want to remove it let me know and I will adapt.

Energy statistics \cite{Szkely2013}
is based on a 
notion of statistical potential energy between probability distributions,
in close analogy to Newton's gravitational potential in classical mechanics. 
It provides a hypothesis test for equality of distributions which is achieved 
under minimum energy. When probability distributions are different the 
statistical potential energy diverges as sample size increases, while tends 
to a nondegenerate limit distribution, i.e. remains finite, when they are 
equal. Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and hierarchical clustering \cite{RizzoClustering} 
by extending Ward's method of minimum variance. Moreover, an application of 
energy statistics to clustering was already proposed \cite{Kgroups}, 
which in part motivated this paper. We refer the reader to \cite{Szkely2013}, 
and references therein, for an overview of energy statistics theory and 
its applications.

In its original formulation, energy statistics has a compact representation
in terms of expectations of pairwise Euclidean distances, providing
straightforward empirical estimates. 
The notion of distance covariance was further generalized from Euclidean 
spaces to metric spaces of strong negative type \cite{Lyons}. Moreover, 
the missing link between energy distance based tests and kernel based tests has 
been recently resolved \cite{Sejdinovic2013}, where a unifying framework
establishing an equivalence between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS), was established. 
This equivalence provides the link between energy distance based measures and 
RKHS based measures, commonly used in machine learning, and form the basis 
of our approach.

The clustering problem has such a long history in machine learning, making it
impossible to mention all important contributions in a short space. 
Perhaps, the most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \cite{Lloyd} of assigning a data point to
the cluster with closest center. The only statistical 
information about each cluster comes from its mean, and it is thus sensitive 
to outliers. Nevertheless, $k$-means works very well when data is 
linearly separable in Euclidean space. Gaussian mixture models (GMM) is also 
very commonly used for clustering, however, it makes strong assumptions about 
the distribution of the data, as $k$-means which is closely related to GMM.

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by using its inner product. Yet, the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually a kernel introduces hyperparameters that 
need to be carefully chosen.
The well-known kernel $k$-means optimization problem is nothing but $k$-means 
in the feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd} of
using the mean of each cluster in the feature space. 
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics was considered
in \cite{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection to kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle behind clustering is to maximize 
the statistical energy,  enforcing probability distributions associated to 
each cluster to be different from one another. We provide a precise 
mathematical formulation to this statement, leading to a quadratically 
constrained quadratic program (QCQP) in the associated RKHS. Our results
immediately establish the connection to kernel methods, by showing that this 
QCQP is equivalent to kernel $k$-means optimization problem. The 
equivalence between kernel $k$-means, spectral clustering, and graph 
partitioning problems is well-known \cite{Dhillon,Dhillon2}. We demonstrate 
how these connections arise from a weighted version of energy statistics.

Our algorithmic contribution is to use Hartigan's method \cite{Hartigan} to 
find local solutions of the above mentioned QCQP, which is NP-hard in general.
Hartigan's method was also used in \cite{Kgroups}, and its advantages
over Lloyd's method was already being demonstrated \cite{Telgarsky,Slonin},
but apparently it did not receive the desired attention. To the 
best of our knowledge, Hartigan's method was not previously applied together 
with kernel methods. We provide an algorithm based on Hartigan's method in 
terms of the Gram matrix of the data computed from a kernel fixed by 
energy statistics. We make clear the advantages of this proposal versus 
Lloyd's method, which kernel $k$-means is based upon and will also be used 
to locally solve our QCQP. We show that both algorithms  have the same
time complexity.

Our numerical results provide compelling evidence that 
Hartigan's method applied to energy statistics based clustering, 
or \emph{energy clustering} for short, is more accurate and robust than 
Lloyd's method. Moreover, in our experiments we put in evidence the
flexibility of energy clustering, which provides a family of default kernels, 
showing that it is able to perform accurately on data coming from 
very different distributions, contrary to $k$-means and GMM for instance.
More specifically, the proposed algorithms for energy clustering perform 
closely to $k$-means and GMM on normally distributed data. On the other
hand, it performs considerably better than $k$-means and GMM on  data that 
is not normally distributed. It also performs better than $k$-means 
and GMM in high dimensions, even on Gaussian settings. However, it performs 
worse than GMM, and more closely to $k$-means, for highly unbalanced clusters
with normal distributions.

Our work is organized as follows. In section~\ref{sec:background}, we introduce
the necessary background on energy statistics and RKHS.
Section~\ref{sec:clustering_theory} contains the main theoretical 
results of this paper,
where we consider a clustering theory based on energy statistics leading
to a QCQP, which is NP-hard in general. We also show the equivalence to 
kernel $k$-means optimization problem.
In Section~\ref{sec:weighted}, we generalize these results to a weighted
version of energy statistics, providing connections to graph
partitioning problems and spectral clustering.
In Section~\ref{sec:twoclass}, we consider the  simple case of a two-class
problem in one dimension,
where we propose an algorithm which requires no random initialization.
In section~\ref{sec:algo}, we consider Lloyd's and Hartigan's methods
and propose iterative algorithms to solve the QCQP.
Section~\ref{sec:numerics} contains some carefully designed numerical
experiments showing that in several settings, energy clustering based
on Hartigan's algorithm
outperms kernel $k$-means algorithm, standard $k$-means algorithm 
and GMM with the expectation maximization algorithm. 
Our final conclusions are presented in Section~\ref{sec:conclusion}.

@@

% @gui: i don't know what this means.  eqn 7 defines something for a population statistic, not a sample statistic.
% @jovo: removed. I was thinking that you can always estimate the expected 
% value by computing a sample mean. but I agree, it's confusing

% @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
% @jovo: I stoped saying non-parametric, however, Kolmogorov-Smirnov is also said to be non-parametric, so I thought that in this context it means that it does not assume any form of the distributions.

@@

The energy distance can be even further generalized.
Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$ and $c_i \in \mathbb{R}$ such that
$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the generalized energy distance
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. This
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a proper metric.

@@

The Moore-Aronszajn theorem 
\cite{Aronszajn}
establishes the converse of the above paragraph. 
For every symmetric
and positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
\mathbb{R}$, there is an associated RKHS $\Hk$ 
with reproducing
kernel $\kk$. The map $\varphi: x \mapsto \kkk_x \in \Hk$ is called
the canonical \emph{feature map}. Given a kernel $\kk$,
this theorem enables us to define an embedding of a probability measure
$P$ into the RKHS as follows: $P \mapsto \kkk_P \in
\Hk$ such that 
$\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
or alternatively $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
We can now  introduce the 
notion of distance between two probability measures using the inner product
of $\Hk$, which is called the maximum mean discrepancy (MMD) and
is given by
\begin{equation}
\label{eq:mmd}
\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk}.
\end{equation}
This can also be written as \cite{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
From the equality between \eqref{eq:mmd} and \eqref{eq:mmd2} we also
have 
\begin{equation}\label{eq:inner_data}
\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y).
\end{equation}
Thus, in practice, we can estimate the inner product between  
embedded distributions 
by averaging the kernel function over sampled data.

The following important result shows that semimetrics of negative
type and symmetric positive definite kernels are closely related
\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type.
We have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
When these conditions are satisfied we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
equivalent kernels.

Now we can state the equivalence between energy distance $\Energy$ and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation} \label{eq:Erho}
\Energy(P, Q) = 
2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] 
= 2 \gamma_\kk^2(P,Q) .
\end{equation}
Due to \eqref{eq:mmd} we can compute the energy distance using the inner 
product of $\Hk$.

Finally, let us recall the main formulas from energy statistics
for the test statistic of equality of distributions \cite{Szkely2013}. 
Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ where
$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
Consider a disjoint partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
$\C_i \cap \C_j = \emptyset$.
Each expectation in the generalized energy distance
\eqref{eq:energy3}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y)
\end{equation}
where $n_i = |\C_i|$ is the number of elements in partition
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv
\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right],
\end{equation}
where $n = \sum_{j=1}^k n_j$.
Given a set of distributions
$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
the quantity $S$ provides
a test statistic for equality of distributions
\cite{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
we have that $S \to \infty$.
This test does not make any assumptions
about the distributions $P_j$.

@@

% @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
% @jovo: ok, will do that after we finish everything. There is a package that automatically remove numbers from unreferenced equations

% @gui: where possible, i recommend using the "name" of the quantity rather than the number of the equation.  eg, here, write S and W, rather than (13) and (12).
% @jovo: ok, I will do that for the main symbols, and refer to equations only when necessary

>>>>

\newcommand\Energy{\mathcal{E}}
\newcommand\EnergyH{\mathcal{E}^{H}}
\newcommand\EnergyL{\mathcal{E}^{L}}
\newcommand\EnergyOne{\mathcal{E}^{1D}}
\newcommand\E{\mathbb{E}}
\newcommand\kk{K}
\newcommand\kkk{h}
\newcommand\Hk{{\mathcal{H}}_{\kk}}
\newcommand\HH{\mathcal{H}}
\newcommand\C{{\mathcal{C}}}
\newcommand\tC{{\widetilde{\C}}}
\newcommand\OO{{\mathcal{O}}}
\newcommand\Zt{Y}
\newcommand{\Ind}[1]{\mathbbm{1}_{#1}}
\newcommand\e{e}
\newcommand\om{\omega}
\newcolumntype{g}{>{\columncolor{gray!20}}l}

@@

\begin{abstract}
Energy statistics was proposed by Sz\' ekely in the 80's inspired by the 
Newtonian gravitational potential from classical mechanics, and it provides 
a non-parametric hypothesis test for equality of distributions. 
It was further generalized from Euclidean spaces to metric spaces of 
strong negative type, and more recently, a connection with reproducing 
kernel Hilbert spaces (RKHS) was established. 
Here we consider the problem of clustering data from an 
energy statistics theory perspective.
We provide a precise mathematical formulation 
yielding a quadratically constrained 
quadratic program (QCQP) in the associated RKHS. We show that this QCQP
is equivalent to kernel $k$-means optimization problem, once the kernel
is fixed.
Thus, our results imply a first principles derivation of kernel $k$-means 
from energy statistics.
However, energy statistics fixes a family of standard kernels.
Furthermore, we also consider a weighted version of energy statistics, 
making connection to graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative algorithm based 
on Hartigan's method, which in this case has the same computational cost 
as kernel $k$-means algorithm, based on Lloyd's method, but usually 
with better clustering quality. 
We provide carefully designed numerical experiments showing the superiority 
of the proposed method compared to kernel $k$-means, standard $k$-means, 
and Gaussian mixture models in a variety of settings.
\end{abstract}

@@

% @jovo: I corrected pretty much all your previous points, and also re-wrote 
% a lot of other stuff throughout the paper. 
% I deleted your previous comments which I believe were taken
% care of, since most of them were simple. The ones which were not, I will
% leave it, in the beggining of each section or in the middle.
%
% @jovo: About ToC, I agree and wouldn't mind remove it, however it is a 
% tradition that most referees like to see. In this case I also included 
% some important information about our work, and it helps to make a transition 
% between introduction and the remaining of the paper. 
% I would leave it, but if you prefer to remove it, let me know and I 
% will adapt the paper.

@@

Energy statistics \cite{Szkely2013}
is based on a 
notion of statistical potential energy between probability distributions,
in close analogy to Newton's gravitational potential in classical mechanics. 
It provides a non-parametric hypothesis test for equality of 
distributions which is achieved 
under minimum energy. When probability distributions are different the 
statistical potential energy diverges as sample size increases, while tends 
to a nondegenerate limit distribution when probability
distributions are equal. 
Energy statistics has been applied to several goodness-of-fit 
hypothesis tests, multi-sample tests of equality of distributions, 
analysis of variance \cite{RizzoVariance}, nonlinear dependence tests through
distance covariance and distance correlation, which generalizes the Pearson
correlation coefficient, and hierarchical clustering \cite{RizzoClustering} 
by extending Ward's method of minimum variance. Moreover, an application of 
energy statistics to clustering was already proposed \cite{Kgroups}, 
which in part motivated this paper. We refer the reader to \cite{Szkely2013}, 
and references therein, for an overview of energy statistics theory and 
its applications.

In its original formulation, energy statistics has a compact representation
in terms of expectations of pairwise Euclidean distances, providing
straightforward empirical estimates. 
The notion of distance covariance was further generalized from Euclidean 
spaces to metric spaces of strong negative type \cite{Lyons}. Furthermore, 
the missing link between energy distance based tests and kernel 
based tests has 
been recently resolved \cite{Sejdinovic2013}, where a unifying framework
establishing an equivalence between generalized energy distances to maximum
mean discrepancies (MMD), which are distances between embeddings of 
distributions in reproducing kernel Hilbert spaces (RKHS), was established. 
This equivalence immediately relates energy statistics to
kernel methods, often used in machine learning, and form the basis 
of our approach.

Clustering has such a long history in machine learning, making it
impossible to mention all important contributions in a short space. 
Perhaps, the most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \cite{Lloyd} of assigning a data point to
the cluster with closest center. The only statistical 
information about each cluster comes from its mean, making it sensitive 
to outliers. Nevertheless, $k$-means works very well when data is 
linearly separable in Euclidean space. Gaussian mixture models (GMM) is 
another very common approach, providing more flexibility than $k$-means, 
however, it still makes strong assumptions about the distribution of 
the data.

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert space by using its inner product. However, the kernel choice remains 
the biggest challenge since there is no principled theory to construct a kernel
for a given dataset, and usually a kernel introduces hyperparameters that 
need to be carefully chosen.
The well-known kernel $k$-means optimization problem is nothing but $k$-means 
in the feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd}
of grouping points that are closer to a cluster center, now
in feature space. 
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics was considered
in \cite{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection with kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the principle behind clustering is to maximize 
the statistical energy,  enforcing probability distributions associated to 
each cluster to be different from one another. We provide a precise 
mathematical formulation to this statement, leading to a quadratically 
constrained quadratic program (QCQP) in the associated RKHS. Our results
immediately establish the connection with kernel methods, showing that this 
QCQP is equivalent to kernel $k$-means optimization problem. The 
equivalence between kernel $k$-means, spectral clustering, and graph 
partitioning problems is well-known \cite{Dhillon,Dhillon2}. We demonstrate 
how these relations arise from a weighted version of energy statistics.

Our main 
algorithmic contribution is to use Hartigan's method \cite{Hartigan} to 
find local solutions of the above mentioned QCQP, which is NP-hard in general.
Hartigan's method was also used in \cite{Kgroups}. More importantly, its 
advantages
over Lloyd's method was already demonstrated 
in some simple settings
\cite{Telgarsky,Slonin}, but apparently this method did not receive 
the deserved attention. To the 
best of our knowledge, Hartigan's method was not previously 
employed together with kernel methods. 
We provide a full kernel based Hartigan's algorithm for clustering,
where the kernel is fixed by energy statistics. 
We make clear the advantages of this proposal versus 
Lloyd's method, which kernel $k$-means is based upon and will also be used 
to solve our QCQP. We show that both algorithms  have the same
time complexity, but Hartigan's method in kernel spaces offer several
advantages. Furthermore, in the examples considered in this paper, it 
also provides superior performance compared to a spectral clustering
relaxation of the QCQP.

Our numerical results provide compelling evidence that 
Hartigan's method applied to energy statistics based clustering, 
or \emph{energy clustering} for short, is more accurate and robust than 
Lloyd's method. Our experiments also put in evidence the
flexibility of energy clustering, which provides a family of default kernels, 
showing that it is able to perform accurately on data coming from 
very different distributions, contrary to $k$-means and GMM for instance.
More specifically, our algorithms for energy clustering perform 
closely to $k$-means and GMM on normally distributed data, but on the other
hand, performs considerably better than $k$-means and GMM on  data that 
is not normally distributed. It also performs better than $k$-means 
and GMM in high dimensions, even on Gaussian settings. However, it performs 
worse than GMM, and more closely to $k$-means, for highly unbalanced clusters
with normal distributions.

Our work is organized as follows. In section~\ref{sec:background}, we introduce
the necessary background on energy statistics and RKHS.
Section~\ref{sec:clustering_theory} contains the main theoretical 
results of this paper,
where we consider a clustering theory based on energy statistics leading
to a QCQP, which is NP-hard in general. We also show the equivalence to 
kernel $k$-means optimization problem.
In Section~\ref{sec:weighted}, we generalize these results to a weighted
version of energy statistics, providing connections to graph
partitioning problems.
In Section~\ref{sec:twoclass}, we consider the  simple case of a two-class
problem in one dimension,
where we propose an algorithm which requires no random initialization.
In section~\ref{sec:algo}, we consider Lloyd's and Hartigan's methods
and propose iterative algorithms to solve the QCQP.
Section~\ref{sec:numerics} contains some carefully designed numerical
experiments showing that, in a variety of settings, energy clustering based
on Hartigan's algorithm
outperform $k$-means, GMM, kernel $k$-means, and a version of
spectral clustering.
Our final conclusions are presented in Section~\ref{sec:conclusion}.

@@

% @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
% @jovo: I stoped saying non-parametric, however, Kolmogorov-Smirnov is also
% said to be non-parametric. I think non-parametric tests also mean that
% it does not assume any form of distributions, i.e. it is a distribution-free
% statistical test.

@@

The energy distance can be even further generalized.
Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$ and $c_i \in \mathbb{R}$ such that
$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the generalized energy distance
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. This
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the associated Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a proper metric.

@@

The Moore-Aronszajn theorem 
\cite{Aronszajn}
establishes the converse of the above paragraph. 
For every symmetric
and positive definite function $\kk: \mathcal{X}\times \mathcal{X} \to
\mathbb{R}$, there is an associated RKHS $\Hk$ 
with reproducing
kernel $\kk$. The map $\varphi: x \mapsto \kkk_x \in \Hk$ is called
the canonical \emph{feature map}. Given a kernel $\kk$,
this theorem enables us to define an embedding of a probability measure
$P$ into the RKHS as follows: $P \mapsto \kkk_P \in
\Hk$ such that 
$\int f(x) d P(x) = \langle f, \kkk_P \rangle$ for all $f \in \Hk$,
or alternatively, $\kkk_P \equiv \int \kk( \, \cdot \,, x)  d P(x)$. 
We can now  introduce the 
notion of distance between two probability measures using the inner product
of $\Hk$, which is called the maximum mean discrepancy (MMD) and
is given by
\begin{equation}
\label{eq:mmd}
\gamma_\kk(P,Q) \equiv \| \kkk_P - \kkk_Q \|_{\Hk}.
\end{equation}
This can also be written as \cite{Gretton2012}
\begin{equation}\label{eq:mmd2}
\gamma_\kk^2(P,Q) = \E \kk(X,X') + \E \kk(Y,Y') - 2 \E \kk(X, Y)
\end{equation}
where $X,X' \stackrel{iid}{\sim} P$ and $Y,Y'\stackrel{iid}{\sim} Q$.
From the equality between \eqref{eq:mmd} and \eqref{eq:mmd2} we also
have 
\begin{equation}\label{eq:inner_data}
\langle \kkk_P, \kkk_Q \rangle_{\Hk} = \E \, \kk(X, Y).
\end{equation}
Thus, in practice, we can estimate the inner product between  
embedded distributions 
by averaging the kernel function over sampled data.

The following important result shows that semimetrics of negative
type and symmetric positive definite kernels are closely related
\cite{Berg1984}. Let $\rho: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$
and $x_0 \in \mathcal{X}$ an arbitrary but fixed point.
Define
\begin{equation}
\label{eq:kernel_semimetric}
\kk(x,y) \equiv 
\tfrac{1}{2} \left[  \rho(x,x_0) + \rho(y,x_0) - \rho(x,y)\right].
\end{equation}
Then, it can be shown that 
$\kk$ is positive definite if and only if $\rho$ is a semimetric
of negative type.
We have a family of kernels, one for each choice of $x_0$. Conversely,
if $\rho$ is a semimetric of negative type and $\kk$ is a kernel in this
family, then 
\begin{equation}
\label{eq:gen_kernel}
\begin{split}
\rho(x,y) &= \kk(x,x) + \kk(y,y) -2\kk(x,y) \\
&=  \| \kkk_x - \kkk_y \|^2_{\Hk}
\end{split}
\end{equation}
and the canonical feature map 
$\varphi: x \mapsto \kkk_x$ is injective \cite{Sejdinovic2013}.
When these conditions are satisfied we say that the kernel $\kk$ 
generates the semimetric $\rho$. 
If two different kernels generate the same $\rho$ they are
said to be equivalent kernels.

Now we can state the equivalence between energy distance $\Energy$ and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation} \label{eq:Erho}
\Energy(P, Q) = 
2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] 
= 2 \gamma_\kk^2(P,Q) .
\end{equation}
Due to \eqref{eq:mmd} we can compute the energy 
distance $\mathcal{E}(P, Q)$ between two probability distributions
using the inner 
product of $\Hk$.

Finally, let us recall the main formulas from energy statistics
for the test statistic of equality of distributions \cite{Szkely2013}. 
Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ where
$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
Consider a disjoint partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
$\C_i \cap \C_j = \emptyset$.
Each expectation in the generalized energy distance
\eqref{eq:energy3}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y) ,
\end{equation}
where $n_i = |\C_i|$ is the number of elements in partition
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j),
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv
\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right],
\end{equation}
where $n = \sum_{j=1}^k n_j$.
Given a set of distributions
$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
the quantity $S$ provides
a test statistic for equality of distributions
\cite{Szkely2013}.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
we have that $S \to \infty$.
Note that this test does not make any assumptions
about the distributions $P_j$, thus it is said to be non-parametric or
distribution-free.

@@

% @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
% @jovo: ok, will do that after we finish everything. There is a package that
% automatically remove numbers from unreferenced equations. It is a common
% practice however to number every equation in the paper,
% specially for reviewing purposes. But later 
% I will remove the numbers from the not cited ones if you don't like
